Examining False Negatives

### 2022-11-01 TestingPipeline/Scripts/examining_false_positives.Rmd
New visualizations of testing data.

---
title: "Viral Sequence Sorting Tools Evaluation"
author: Bridget Hegarty, James Riddell
date: 07-22-2022
output: html_notebook
---
This Rmarkdown file assesses the output of CheckV, DeepVirFinder, Kaiju,
VIBRANT, VirSorter, and VirSorter2 on multiple training sets of microbial DNA, 
primarily from NCBI. Created from fungal, viral, bacterial, archeael, protist,
and plasmid DNA sequences

@@

# Assessing performance against the "truth"
note that this is only as accurate as the annotations of the input sequences

this function calculates the precision, recall, and F1 score for each pipeline
```{r}
assess_performance <- function(seqtype, keep_score) {

  truepositive <- rep("not viral", length(seqtype))
  truepositive[seqtype=="virus"] <- "viral"

  #make confusion matrix
  confusion_matrix <- rep("true negative", length(keep_score))
  confusion_matrix[truepositive=="viral" & keep_score<=1] <- "false negative"
  confusion_matrix[truepositive=="viral" & keep_score>=1] <- "true positive"
  confusion_matrix[truepositive=="not viral" & keep_score>=1] <- "false positive"

  TP <- table(confusion_matrix)[4]
  FP <- table(confusion_matrix)[2]
  TN <- table(confusion_matrix)[3]
  FN <- table(confusion_matrix)[1]

  precision <- TP/(TP+FP)
  recall <- TP/(TP+FN)
  F1 <- 2*precision*recall/(precision+recall)

  MCC <- (TP*TN-FP*FN)/sqrt(as.numeric(TP+FP)*as.numeric(TP+FN)*as.numeric(TN+FP)*as.numeric(TN+FN))

  auc <- round(auc(truepositive, keep_score),4)

  #by type metrics
  fungal_FP <- table(confusion_matrix[seqtype=="fungi"])[2]
  protist_FP <- table(confusion_matrix[seqtype=="protist"])[2]
  bacterial_FP <- table(confusion_matrix[seqtype=="bacteria"])[2]
  viral_FN <- table(confusion_matrix[seqtype=="virus"])[1]

  performance <- c(precision, recall, F1, MCC, auc, fungal_FP, 
                   protist_FP, bacterial_FP, viral_FN)
  names(performance) <- c("precision", "recall", "F1", "MCC", "AUC", "fungal_FP",
                          "protist_FP", "bacterial_FP", "viral_FN")

  return(performance)
}
```

combination of tools list
```{r}
combos_list <- data.frame(toolcombo=rep(0, 64),
                          tune_not_viral=rep(0, 64),
                          DVF=rep(0, 64),
                          tune_viral=rep(0, 64),
                          VIBRANT=rep(0, 64),
                          VS=rep(0, 64),
                          VS2=rep(0, 64))
p <- 1

for (i in c(0,1)){
  for (j in c(0,1)){
    for (k in c(0,1)){
      for (l in c(0,1)){
        for (m in c(0,1)){
          for (n in c(0,1)){
            combos_list$toolcombo[p] <- paste(i,j,k,l,m,n)
            combos_list$toolcombo2[p] <- paste(if(i){"tv"}else{"0"},if(j){"DVF"}else{"0"},
                                               if(k){"tnv"}else{"0"},if(l){"VB"}else{"0"},
                                               if(m){"VS"}else{"0"},if(n){"VS2"}else{"0"})
            combos_list$tune_not_viral[p] <- i
            combos_list$DVF[p] <- j
            combos_list$tune_viral[p] <- k
            combos_list$VIBRANT[p] <- l
            combos_list$VS[p] <- m
            combos_list$VS2[p] <- n
            p <- p+1
          }
        }
      }
    }
  }
}

combos_list <- combos_list[-1,]
```

this function builds a list of all of the combinations that the user wants to 
test. 
In this case, we're comparing the performance of all unique combinations of the 
six tools.
```{r}
build_score_list <- function(input_seqs, combos) {
  output <- data.frame(precision=rep(0, nrow(combos)),
                       recall=rep(0, nrow(combos)),
                       F1=rep(0, nrow(combos)),
                       MCC=rep(0, nrow(combos)),
                       AUC=rep(0, nrow(combos)),
                       fungal_FP=rep(0, nrow(combos)),
                       protist_FP=rep(0, nrow(combos)),
                       bacterial_FP=rep(0, nrow(combos)),
                       viral_FN=rep(0, nrow(combos)))
  for (i in 1:nrow(combos)) {
    keep_score <- getting_viral_set_1(input_seqs, include_vibrant = combos$VIBRANT[i],
                                            include_virsorter = combos$VS[i],
                                            include_virsorter2 = combos$VS2[i],
                                            include_tuning_viral = combos$tune_viral[i],
                                            include_tuning_not_viral = combos$tune_not_viral[i],
                                            include_deepvirfinder = combos$DVF[i])

    output[i,1:9] <- assess_performance(input_seqs$seqtype, keep_score)

    output$toolcombo[i] <- paste(combos$tune_viral[i],combos$DVF[i],
                                 combos$tune_not_viral[i], combos$VIBRANT[i],
                                 combos$VS[i], combos$VS2[i])
  }

  output[is.na(output)] <- 0

  return (output)
}
```

## Calculate the performance of each pipeline
```{r}
accuracy_scores <- data.frame(testing_set_index=rep(0, nrow(combos_list)*10),
                      precision=rep(0, nrow(combos_list)*10),
                       recall=rep(0, nrow(combos_list)*10),
                       F1=rep(0, nrow(combos_list)*10),
                       MCC=rep(0, nrow(combos_list)*10), 
                      AUC=rep(0, nrow(combos_list)*10),
                      fungal_FP=rep(0, nrow(combos_list)*10),
                      protist_FP=rep(0, nrow(combos_list)*10),
                      bacterial_FP=rep(0, nrow(combos_list)*10),
                      viral_FN=rep(0, nrow(combos_list)*10))

accuracy_scores <- cbind(testing_set_index=rep(1, nrow(combos_list)),
                              build_score_list(viruses[viruses$Index==1,], combos_list))
for (i in 2:10) {
  accuracy_scores <- rbind(accuracy_scores,
                           cbind(testing_set_index=rep(i, nrow(combos_list)),
                              build_score_list(viruses[viruses$Index==i,], combos_list)))
}
```

```{r}
library("stringr")
```

```{r}
accuracy_scores$numtools <- str_count(accuracy_scores$toolcombo, "1")
#accuracy_scores <- accuracy_scores[order(accuracy_scores$numtools, decreasing=F),]
accuracy_scores <- accuracy_scores[order(accuracy_scores$MCC, decreasing=F),]
accuracy_scores$toolcombo <- factor(accuracy_scores$toolcombo, levels = unique(accuracy_scores$toolcombo))
accuracy_scores$numtools <- as.factor(accuracy_scores$numtools)
```

## Visualize how the precision, recall, and F1 scores change across pipelines.
```{r}
pal <- ggthemes::tableau_color_pal(palette="Tableau 10", type="regular")
p2 <- ggplot(accuracy_scores, aes(x=toolcombo, y=F1, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("F1 Score")
p2
ggplot(accuracy_scores, aes(x=toolcombo, y=precision, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Precision")
ggplot(accuracy_scores, aes(x=toolcombo, y=recall, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Recall")
ggplot(accuracy_scores, aes(x=precision, y=recall, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Precision") +
  ylab("Recall")
ggplot(accuracy_scores, aes(x=toolcombo, y=abs(precision-recall), 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Precision-Recall")
ggplot(accuracy_scores, aes(x=toolcombo, y=MCC, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("MCC")
ggplot(accuracy_scores, aes(x=toolcombo, y=AUC, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("AUC")
ggplot(accuracy_scores, aes(x=toolcombo, y=fungal_FP, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Fungal False Positives")

ggplot(accuracy_scores, aes(x=toolcombo, y=protist_FP, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Protist False Positives")

ggplot(accuracy_scores, aes(x=toolcombo, y=bacterial_FP, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Bacterial False Positives")
ggplot(accuracy_scores, aes(x=toolcombo, y=viral_FN, 
                                  color=numtools, fill=numtools)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  xlab("Tool Combination (tv, DVF, tnv, VB, VS, VS2)") +
  ylab("Viral False Negatives")
```

```{r}
accuracy_scores_melt <- accuracy_scores %>% 
  select(testing_set_index, precision, recall, MCC, numtools, toolcombo) %>%
  pivot_longer(cols=c(precision, recall, MCC), 
               names_to="performance_metric",
               values_to="performance_metric_score")
```

```{r}
ggplot(accuracy_scores_melt, aes(x=numtools, y=performance_metric_score, 
                                  color=numtools, fill=numtools)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

```
comparing metric with and without tuning rules
```{r}
accuracy_scores_melt$tuning_inc <- "no"
accuracy_scores_melt$tuning_inc[substring(accuracy_scores_melt$toolcombo, 1, 1)==1] <- "tv"
accuracy_scores_melt$tuning_inc[substring(accuracy_scores_melt$toolcombo, 3, 3)==1] <- "tnv"
accuracy_scores_melt$tuning_inc[substring(accuracy_scores_melt$toolcombo, 1, 1)==1 &
                                  substring(accuracy_scores_melt$toolcombo, 3, 3)==1] <- "both"
```

```{r}
ggplot(accuracy_scores_melt, aes(x=tuning_inc, y=performance_metric_score)) +
  geom_boxplot() +
  geom_point(aes(color=numtools, fill=numtools), alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

ggplot(accuracy_scores_melt, aes(x=tuning_inc, y=performance_metric_score,
                                 color=numtools, fill=numtools)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

ggplot(accuracy_scores_melt, aes(x=tuning_inc, y=performance_metric_score)) +
  geom_boxplot() +
  geom_boxplot(aes(color=numtools, fill=numtools)) +
 # geom_point(aes(color=numtools, fill=numtools), alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14, angle = 90),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("MCC") +
  xlab("Number of Tools") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  facet_wrap(~performance_metric)

```

```{r}
write_tsv(accuracy_scores, "20221029_accuracy_scores.tsv")
```

to do: add in clustering and ordination like in the drinking water R notebook

# Experimenting

## high precision example
```{r}
viruses$keep_score_high_precision <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = F,
                                              include_tuning_viral = F,
                                              include_tuning_not_viral = T,
                                              include_virsorter = F)
```

```{r}
viruses$confusion_matrix_high_precision <- "true negative"
viruses$confusion_matrix_high_precision[viruses$seqtype=="virus" & viruses$keep_score_high_precision<1] <- "false negative"
viruses$confusion_matrix_high_precision[viruses$seqtype=="virus" & viruses$keep_score_high_precision>=1] <- "true positive"
viruses$confusion_matrix_high_precision[viruses$seqtype!="virus" & viruses$keep_score_high_precision>=1] <- "false positive"
```

visualizing confusion matrix by taxa
```{r}
confusion_by_taxa <- melt(table(viruses$confusion_matrix_high_precision, viruses$seqtype, viruses$Index))
colnames(confusion_by_taxa) <- c("confusion_matrix", "seqtype","Index", "count")
```

```{r}
length(grep("true", viruses$confusion_matrix_high_precision))/nrow(viruses)
```

@@

```{r}
ggplot(confusion_by_taxa, aes(x=count, y=as.factor(Index),
                   fill=confusion_matrix,
                   color=confusion_matrix)) +
  geom_bar(stat="identity") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(4)), 0.5),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(4)), 1),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  xlab("Number of Sequences") +
  ylab("") + 
  facet_wrap(~seqtype, scales = "free") +
  coord_flip()
```
this rule set had the highest precision, but as you can see, this comes with a big sacrifice in recall

## high MCC example
```{r}
viruses$keep_score_high_MCC <- getting_viral_set_1(viruses, include_deepvirfinder = F,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_tuning_viral = T,
                                              include_tuning_not_viral = T,
                                              include_virsorter = T)
```

accuracy:
```{r}
length(grep("true", viruses$confusion_matrix_high_MCC))/nrow(viruses)
```

recall
```{r}
length(grep("true positive", viruses$confusion_matrix_high_MCC))/length(grep("virus", viruses$seqtype))
```

visualizing confusion matrix by taxa
```{r}
confusion_by_taxa <- viruses %>% count(confusion_matrix_high_MCC, seqtype, Index)

colnames(confusion_by_taxa) <- c("confusion_matrix", "seqtype","index", "count")
```

```{r}
ggplot(confusion_by_taxa, aes(x=count, y=as.factor(index),
                   fill=confusion_matrix,
                   color=confusion_matrix)) +
  geom_bar(stat="identity") +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(4)), 0.5),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(4)), 1),
                    labels=c("false negative", "false positive", 
                             "true negative", "true positive")) +
  xlab("Number of Sequences") +
  ylab("") + 
  facet_wrap(~seqtype, scales = "free") +
  coord_flip()
```

differences based on genome size
```{r}
viruses$size_class <- "3-5kb"
viruses$size_class[viruses$checkv_length>5000] <- "5-10kb"
viruses$size_class[viruses$checkv_length>10000] <- ">10kb"
```

```{r}
confusion_by_taxa <- viruses %>% count(confusion_matrix_high_MCC, seqtype, size_class, Index)

colnames(confusion_by_taxa) <- c("confusion_matrix", "seqtype","size", "index", "count")
```

```{r}
confusion_vir_called <- confusion_by_taxa %>% filter(confusion_matrix=="true positive" | confusion_matrix=="false positive")

type_count <- viruses %>% count(seqtype, size_class, Index)

confusion_vir_called$per_viral <- 0

for (i in c(1:nrow(confusion_vir_called))) {
  confusion_vir_called$per_viral[i] <- confusion_vir_called$count[i]/type_count$n[type_count$seqtype==confusion_vir_called$seqtype[i] & 
                                                                                    type_count$Index==confusion_vir_called$index[i] &
                                                                                    type_count$size_class==confusion_vir_called$size[i]]*100
}

confusion_vir_called <- confusion_vir_called %>% group_by(seqtype, size) %>%
  summarise(mean=mean(per_viral), 
            sd=sd(per_viral))

confusion_vir_called$size <- factor(confusion_vir_called$size,
                                    levels = c("3-5kb", "5-10kb", ">10kb"))
```

```{r}
ggplot(confusion_vir_called, aes(y=mean, x=size,
                   fill=seqtype,
                   color=seqtype)) +
  geom_bar(stat="identity", position=position_dodge()) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2,
                 position=position_dodge(.9)) +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(6)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(6)), 1)) +
  xlab("Length") +
  ylab("Sequences Called Viral (%)") 
```

```{r}
viruses$keep_score_vb <- getting_viral_set_1(viruses, include_deepvirfinder = F,
                                              include_vibrant = T,
                                              include_virsorter2 = F,
                                              include_tuning_viral = F,
                                              include_tuning_not_viral = F,
                                              include_virsorter = F)

viruses$keep_score_vb_dvf <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = F,
                                              include_tuning_viral = F,
                                              include_tuning_not_viral = F,
                                              include_virsorter = F)

viruses$keep_score_vb_dvf_vs2 <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_tuning_viral = F,
                                              include_tuning_not_viral = F,
                                              include_virsorter = F)

viruses$keep_score_vb_dvf_vs2_vs <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_tuning_viral = F,
                                              include_tuning_not_viral = F,
                                              include_virsorter = T)

viruses$keep_score_vb_dvf_vs2_vs_tv <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_tuning_viral = T,
                                              include_tuning_not_viral = F,
                                              include_virsorter = T)

viruses$keep_score_vb_dvf_vs2_vs_tv_tnv <- getting_viral_set_1(viruses, include_deepvirfinder = T,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_tuning_viral = T,
                                              include_tuning_not_viral = T,
                                              include_virsorter = T)
```

# Considering how each method contributes to the final prediction
```{r}
viruses_high <- viruses[viruses$keep_score_vb_dvf_vs2_vs_tv>=1,] 
viruses_high_mod <- viruses_high %>% select(keep_score_vb,keep_score_vb_dvf, 
                                            keep_score_vb_dvf_vs2, keep_score_vb_dvf_vs2_vs, 
                                            keep_score_vb_dvf_vs2_vs_tv, keep_score_vb_dvf_vs2_vs_tv_tnv)
#viruses_high_mod <- apply(viruses_high_mod, c(1,2), function(x) {if (x >= 1) {x <- 1} else {x <- 0}})
viruses_high_mod <- as_tibble(viruses_high_mod)

```

```{r}
sm_m <- reshape2::melt(viruses_high_mod)
colnames(sm_m) <- c("method", "viral_score")
```

```{r}
sm_m <- sm_m[sm_m$viral_score>0,]

sm_m$score <- sm_m$viral_score

sm_m$score[sm_m$viral_score==0.5] <- "0.5"
sm_m$score[sm_m$viral_score>=1] <- "1"
sm_m$score[sm_m$viral_score>=2] <- "2"
sm_m$score[sm_m$viral_score>=3] <- "3"
sm_m$score[sm_m$viral_score>=4] <- "4"
sm_m$score[sm_m$viral_score>=5] <- "5"

sm_m$score <- factor(sm_m$score, 
                                       levels=c("0.5", "1", "2","3","4","5"))
```

```{r}
ggplot(sm_m, aes(x=method, y=score,
                   fill=score)) +
  geom_bar(stat="identity") +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    legend.position = "bottom"
  ) +
  scale_fill_manual(name = 'Viral Score',
                     values = alpha(c(viridis(6)), 1)) +
  xlab("") +
  ylab("Number of Sequences") +
  coord_flip()
```
## Visualizing confusion matrix by number of tools

```{r}
viruses$keep_score_visualize <- viruses$keep_score_high_MCC
viruses$keep_score_visualize[viruses$keep_score_high_MCC>1] <- "> 1"
viruses$keep_score_visualize[viruses$keep_score_high_MCC==1] <- "1"
viruses$keep_score_visualize[viruses$keep_score_high_MCC==0.5] <- "0.5"
viruses$keep_score_visualize[viruses$keep_score_high_MCC==0] <- "0"
viruses$keep_score_visualize[viruses$keep_score_high_MCC==-0.5] <- "-0.5"
viruses$keep_score_visualize[viruses$keep_score_high_MCC==-1] <- "-1"
viruses$keep_score_visualize[viruses$keep_score_high_MCC<=-1] <- "< -1"

viruses$keep_score_visualize <- factor(viruses$keep_score_visualize, 
                                       levels=c("< -1", "-1", "-0.5", "0", "0.5","1", "> 1"))
#viruses$keep_score_visualize <- factor(viruses$keep_score_visualize, 
#                                       labels=c("≤ 0", "≤ 0", "≤ 0", "0.5","1", "> 1"))
```

```{r}
levels(factor(viruses$keep_score_visualize))
```

>>>>

---
title: "Examining False Negatives"
author: Bridget Hegarty, James Riddell
date: 10-30-2022
output: html_notebook
---
This Rmarkdown file examines the false positives from the output of CheckV, DeepVirFinder, Kaiju,
VIBRANT, VirSorter, and VirSorter2 on multiple training sets of microbial DNA, 
primarily from NCBI. Created from fungal, viral, bacterial, archeael, protist,
and plasmid DNA sequences

@@

```{r}
viruses$keep_score_high_MCC <- getting_viral_set_1(viruses, include_deepvirfinder = F,
                                              include_vibrant = T,
                                              include_virsorter2 = T,
                                              include_tuning_viral = T,
                                              include_tuning_not_viral = T,
                                              include_virsorter = T)
```

getting a list of false positive bacteria and plasmids from testing set 1
```{r}
fp_bacteria_list <- viruses$contig[viruses$Index==1 & 
                                     viruses$seqtype=="bacteria" &
                                     viruses$confusion_matrix_high_MCC=="false positive"]

fp_plasmid_list <- viruses$contig[viruses$Index==1 & 
                                     viruses$seqtype=="plasmid" &
                                     viruses$confusion_matrix_high_MCC=="false positive"]

tp_virus_list <- viruses$contig[viruses$Index==1 & 
                                     viruses$seqtype=="virus" &
                                     viruses$confusion_matrix_high_MCC=="true positive"]
```

vibrant annotations
```{r}
annotations_vb <- read_tsv("../ToolOutput/VIBRANT_annotations_metagenomic_testing_set_1.tsv")
```
```{r}
annotations_vb <- separate(annotations_vb, col="scaffold", into=c("seqtype", "contig"),
                           sep="--")
annotations_vb$contig <- sub("\\.", "_", annotations_vb$contig)
```

```{r}
head(annotations_vb)
```
```{r}
table(annotations_vb$contig %in% fp_bacteria_list)
table(annotations_vb$contig %in% fp_plasmid_list)
```

```{r}
annotations_vb_fp <- rbind(annotations_vb[annotations_vb$contig %in% fp_bacteria_list,],
                           annotations_vb[annotations_vb$contig %in% fp_plasmid_list,],
                           annotations_vb[annotations_vb$contig %in% tp_virus_list,])
```

```{r}
annotations_vb_fp_vog_count <- annotations_vb_fp %>%
  select(contig, VOG, seqtype) %>%
  group_by(contig, seqtype) %>%
  summarize(num_vog=table(is.na(VOG))[[1]],
            prop_vog=table(is.na(VOG))[[1]]/length(VOG),
            num_orfs=length(VOG))
```

```{r}
annotations_vb_fp_vog_count$confusion_matrix <- "false positive"
annotations_vb_fp_vog_count$confusion_matrix[annotations_vb_fp_vog_count$seqtype=="virus"] <- "true positive"
```

@@

```{r}
ggplot(annotations_vb_fp_vog_count, 
       aes(x=seqtype, y=prop_vog, 
           color=seqtype, fill=seqtype)) +
  geom_boxplot() +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("Proportion Genes VOGs") +
  xlab("Sequence Type") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(3)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(3)), 1)) +
  facet_wrap(~confusion_matrix, scales="free_x")
```
```{r}
t.test(annotations_vb_fp_vog_count$prop_vog[annotations_vb_fp_vog_count$seqtype=="bacteria"],
       annotations_vb_fp_vog_count$prop_vog[annotations_vb_fp_vog_count$seqtype=="virus"])
```
```{r}
t.test(annotations_vb_fp_vog_count$prop_vog[annotations_vb_fp_vog_count$seqtype=="plasmid"],
       annotations_vb_fp_vog_count$prop_vog[annotations_vb_fp_vog_count$seqtype=="virus"])
```

```{r}
ggplot(annotations_vb_fp_vog_count, 
       aes(x=num_vog, y=prop_vog, 
           color=seqtype, fill=seqtype)) +
  geom_point(alpha=0.5) +
  theme_light() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    axis.text.y=element_text(size=14),
    axis.text.x=element_text(size=14),
    legend.text=element_text(size=12),
    axis.title=element_text(size=16),
  ) +
  ylab("Proportion Genes VOGs") +
  xlab("Number of VOGs") +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(3)), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(3)), 1))
```

# Thinking about short protists
```{r}
viruses[viruses$confusion_matrix_high_MCC=="false positive" & viruses$seqtype=="protist",]

table(viruses$confusion_matrix_high_MCC[viruses$seqtype=="protist"], 
      viruses$max_score_group[viruses$seqtype=="protist"])

table(viruses$Kingdom[viruses$seqtype=="protist" &  viruses$confusion_matrix_high_MCC=="false positive"], 
      viruses$max_score_group[viruses$seqtype=="protist" & viruses$confusion_matrix_high_MCC=="false positive"])

table(viruses$percent_viral[viruses$seqtype=="protist" &  viruses$confusion_matrix_high_MCC=="false positive"]>20, 
      viruses$max_score_group[viruses$seqtype=="protist" & viruses$confusion_matrix_high_MCC=="false positive"])
```

```{r}
ggplot(viruses, aes(y=percent_viral, x=confusion_matrix_high_MCC)) +
  geom_boxplot() +
  geom_point() +
  theme_bw() +
  facet_wrap(~seqtype, scales = "free") 
```
```{r}
table(viruses$confusion_matrix_high_MCC[viruses$checkv_length>=5000])
```
